Current state and prospects of R-packages for the design of experiments

### 2022-06-13 paper.Rmd
rechecked up to Section 3.1 (inclusive)

---
title: Current state and prospects of R-packages for the design of experiments
author:
  - name: Emi Tanaka
    affiliation: Monash University
  - name: Dewi Amaliah 
    affiliation: Monash University
abstract: >
    Re-running an experiment is generally costly and in some cases impossible due to limited resources, so the design of an experiment plays a critical role in increasing the quality of experimental data.  In this article we describe the current state of the R-packages for the design of experiments through an exploratory data analysis of package downloads, package metadata, and the comparison of characteristics with other topics. We observe that experimental designs in practice appear to be sufficiently manufactured by a small number of packages and the development of experimental designs occur in silos. We discuss also the interface designs of widely utilised R packages in the field of experimental design and discuss its future prospects for the field.
bibliography: paper.bib
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
keywords: ["experimental design", "CRAN task view", "user interface", "application programming interface", "software design"]
output: 
  #bookdown::html_document2: 
  #  theme: paper
  # below is to have a JSS paper style template
  pagedown::html_paged:
    template: assets/html-template.html
    csl: assets/html-template.csl
    self_contained: false
    css: ["assets/html-template.css"]
---

```{css, echo = FALSE}
h1, h2, h3, div > * {
 font-family: "Helvetica Neue",Helvetica,Arial,sans-serif;
}
```

```{r setup, cache = FALSE, include = FALSE}
library(targets)
library(glue)
library(tidyverse)
library(patchwork)
library(tidygraph)
library(ggraph)
source("R/plot.R")
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.path = "figures/",
                      fig.width = 6,
                      fig.height = 6.5,
                      message = FALSE,
                      warning = FALSE)

@@

tar_load(npkgs)
tar_load(gini_yearly)
tar_load(gini_yearly_ctv)
```

@@

The critical role of data collection is well captured in the expression "garbage in, garbage out" -- in other words, if the collected data is rubbish then no analysis, however complex it may be, can make something out of it. A carefully crafted data collection scheme is therefore critical to optimise the information from data. Of all the data collection schemes, the field of experimental designs is specifically devoted to planning an experiment largely based on the founding principles by @Fisher1935-qc or an optimisation framework like those described in @Pukelsheim2006-sv. These experimental designs are often constructed with the aid of a statistical software, such as R [@R], so software usage can aid in understanding the practice of experimental design. Given the large gain that can be made from collecting higher quality experimental data, we benefit from understanding the current state of the experimental designs in practice.

@@

Methods for data collection can be dichotomised by the type of data collected -- namely, experimental or observational -- or alternatively, categorised as experimental design (including quasi-experimental design) or survey design. This dichotomisation, to a great extent, is seen in the [Comprehensive R Archive Network (CRAN) task views](https://cran.r-project.org/web/views/) (a volunteer maintained list of R-packages by topic) where R-packages in experimental design are in `r ctv("ExperimentalDesign")` task view and R-packages in survey designs are in `r ctv("OfficialStatistics")` task view. The full list of available topics can be see in Table S1 in the Supplementary Materials and also accessed via the `ctv` package [@ctv]. A subset of experimental designs are segregated into `r ctv("ClinicalTrials")` task view, where the focus is on clinical trials with primary interest in sample size calculations. This paper focuses on the packages in `r ctv("ExperimentalDesign")` task view, henceforth referred to as "DoE packages".

@@

A utility of software can also be described by its design to facilitate clear expression and interpretation of the desired experimental design. Certain programming language design can hinder or discourage development of a reliable program [@Wasserman1975-xr].  The immense popularity of `tidyverse` [a collection of R-packages for various stages of data analysis that place enormous emphasis on the interface design by @Wickham2019-mj] is a testament to the impact an interface design can have in practice.

The paper is organised as follows. Section `r ref("data")` briefly describes the data source used for the analysis in Section `r ref("eda")`; Section `r ref("eda")` presents some insights into the state of the current DoE packages by exploratory data analysis of package download data, text descriptions and comparisons with other CRAN task views; Section `r ref("design")` discuss the interface designs of DoE packages, and we conclude with a discussion in Section `r ref("discussion")` of future prospects.

@@

To study the DoE packages, we analyse data using three sources of data as described next.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers located across the world that store mirrored versions of R and R-packages. The most popular network is the RStudio mirror (the default server for those that use the RStudio IDE). The RStudio mirror is also the only server that provides a comprehensive daily download logs of R and R-packages since October 2012. The summary data can be easily accessed with the `cranlogs` package [@cranlogs]. This paper uses the data from the beginning of 2013 to end of 2021 (a total of 9 years) for the DoE packages.

@@

The list of packages in each CRAN task views (as of `r as.Date(file.info("_targets/objects/ctv_pkgs_data")$mtime)`) are used to contrast the characteristics of DoE packages in Section `r ref("silo")`.

@@

All results presented are derived from exploratory data analysis of observational data; consequently, all interpretations are somewhat speculative and 
may not be indicative of the true state of the field of experimental design. In particular, any analysis over time are confounded by the fact that the nature of users and package management have also changed over the years. It should be noted that some DoE packages may have been archived or removed from the task view over the years so any cross-sectional analysis presented may not reflect the set of all DoE packages at that particular time period (although we assume such incidences are low).

A subset of DoE packages is not primarily about design of experiments but about the analysis of experimental data. A complete delineation of these packages is difficult as many have some functions that can aid decisions or constructions of experimental designs (and any categorisation is prone to our subjective bias) so we opted not to remove any DoE packages in the analysis.

## Small, but diverse, set of packages are sufficient for most designs {#popular}

@@

```{r plot-lorenz, fig.height = 4, fig.width = 4, fig.cap = "Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality."}
tar_read(plot_lorenz2021) +
  plot_setup() +
  labs(y = "Cumulative share of downloads (%)",
       x = "Percentage of DoE packages") +
  theme(plot.margin = margin(r = 20, t = 20)) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10))
```

 If we consider package downloads as a measure of "wealth", then we can consider using the Gini index [@Gini1921-mf] as a measure of download inequality across packages. The ratio of the red region over the total colored regions in Figure `r ref("fig:plot-lorenz")` corresponds to the Gini index for 2021. A Gini index of 0\% indicates equality in downloads across packages while a value of 100\% indicates maximal inequality (all downloads are due to one package). In Figure `r ref("fig:download-share")`, we see that the distributions of the downloads each year have a heavy right tail with the Gini index ranging from `r scales::percent(min(gini_yearly$gini), 0.1)` to `r scales::percent(max(gini_yearly$gini), 0.1)` across the years 2013 to 2021, indicating that there is a high level of inequality of downloads across packages, particularly with more pronounced inequality in the last 5 years.

```{r download-share, fig.width = 5.5, fig.height = 4, fig.cap = "The distribution of the number of downloads for each DoE package by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 5 years, the Gini index is consistent above 60\\% each year indicating that most downloads are due to a relatively small number of packages."}
tar_read(plot_download_dist)
```

While in absolute terms the Gini index is high for `r ctv("ExperimentalDesign")`, the inequality is not as severe as other CRAN task views as shown in Figure `r ref("fig:fig-gini-all-ctvs")`. We can see in Figure `r ref("fig:fig-gini-all-ctvs")` that the Gini index is generally increasing for DoE packages (as is generally the case for other CRAN task views as shown in Figure S1 in the Supplementary Materials).

```{r fig-gini-all-ctvs, fig.height = 4, fig.width = 4, fig.cap = "The points show the Gini index of the download counts by year for ExperimentalDesign CRAN task view with the color showing the number of packages. The grey line shows the distribution of the Gini index across years for all other CRAN task views. See Figure S2 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view."}
gini_yearly_ctv %>% 
  ungroup() %>%  
  arrange(ctv, year) %>% 
  mutate(ctv = fct_reorder(ctv, gini, last),
         year = factor(year),
         gini = 100 * gini) %>%  
  ggplot(aes(year, gini, group = ctv)) + 
  geom_line(color = "grey") +
  geom_line(data = ~filter(.x, ctv=="ExperimentalDesign")) + 
  geom_point(data = ~filter(.x, ctv=="ExperimentalDesign"),
             aes(color = n))  +
  plot_setup() +
  labs(x = "Year", y = "Gini index (%)", color = "# of packages",
       title = "ExperimentalDesign CRAN task view") +
  guides(color = guide_colorbar(direction = "horizontal")) +
  theme(legend.position = "bottom") +
  scale_colour_distiller(palette = "YlOrRd", direction = 1) 
```

From Figures `r ref("fig:download-share")` and `r ref("fig:fig-gini-all-ctvs")`, we could draw some indicative states of the field of experimental designs with possible counterfactual interpretations:

* The increase in the number of packages that are not highly downloaded may mean that **_there are more packages to construct niche experimental designs_**. Some examples of these packages include `qtlDesign`, `PwrGSD` and `Crossover` made for QTL experiments, group sequential designs and crossover trials, respectively. These packages would naturally have a smaller number of potential users. Counterfactual to this, the increase could be due to other external factors, such as an increase in a number of skilled contributors, a change in CRAN policy or management to add packages (either to CRAN and/or task view), and/or that new packages are still yet to amass users. While there is an argument that low download counts are due to the low utility and/or quality of the packages, packages in CRAN task views are selected by expert maintainers; thus we can reasonably assume that any package listed in CRAN task view are of a decent utility and quality. 
* If the downloads are reflective of the experimental designs used in practice, **_small set of packages appear to be sufficient for most to construct the full set of designs of experiments needed in practice_**. Packages of course evolve and the top downloaded packages have had regular updates that may have broadened its scope. 
* While small subset of DoE packages are most frequently used, none of these packages are dominant in the field (judging from comparisons with other CRAN task views). This suggests that there are **_diverse approaches to designing experiments_**. This observation doesn't take into account other approaches to generating experimental designs, like the proprietary software, CycDesignN [@cycdesignn] and SAS [@sas1985sas], which may be widely used.

## There is a lack of adaptation of new or innovative designs  {#ranking}

@@

```{r rank-over-time, fig.width = 5.5, fig.height = 4.5, fig.cap = "The plot shows the rank of top 10 downloaded packages by year. "}
tar_read(plot_rank_dist)
```

@@

```{r established-packages, fig.height = 5, fig.cap = "The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log, with base 10, of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R."}
tar_read(cran_rank_yearly) %>%
  mutate(nupdates = map2_int(package, year, ~{
    tar_read(updates_all) %>% 
      filter(package == .x,
             lubridate::year(update) <= .y) %>% 
      nrow()
  })) %>% 
  ggplot(aes(first, total/1000)) +
  geom_point(aes(color = nupdates)) + 
  geom_label(data = ~ .x %>% 
              group_by(year) %>% 
              summarise(R = cor(as.numeric(first), log10(total/1000)) %>% 
                          scales::comma(0.01) %>% 
                          paste0("r = ", .),
                        y = quantile(total/1000, 0.99)),
            aes(label = R, y = y, x = as.Date("2014-11-01")),
            size = 3) +
  facet_wrap(~year, scale = "free_y") +
  geom_smooth(se = FALSE, 
              method = lm, 
              formula = y ~ x) +
  scale_y_log10() + 
  scale_color_continuous(trans = "log10") + 
  plot_setup() +
  labs(x = "First release date",
       y = "Total download (1000s)",
       color = "# of updates") +
  theme(legend.position = "bottom")
```

@@

```{r wordcloud-over-time, fig.width = 5.5, fig.height = 5, fig.cap = "The above figure shows the word cloud of bigrams from the title and descriptions of the CRAN packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram."}
tar_read(plot_wordcloud2) + 
  scale_size_area(max_size = 5) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 12)) +
  labs(color = "Total downloads (%)")
```

@@

In this section, we discuss these interface designs broadly, with examples illustrated from the top downloaded packages (shown in Figure `r ref("fig:rank-over-time")`). We exclude `ez` and `DiceKriging` from the discussion as the former  is predominately visualisation of experimental data and latter is about the analysis of computer experiments in addition to belonging to the same suite of packages as `DiceDesign`.

`DoE.wrapper` [@DoE.wrapper]

@@

Computer experiments, as discussed in Section `r ref("topics")`, which generally involve space filling designs, have been on the rise. The exemplar for this type of designs are `lhs` and `DiceDesign`. For the `lhs` package, functions like `randomLHS()`, `optimumLHS()`, and `maximinLHS()`, require users to specify the sample size ($n$) and the number of variables ($p$) then it generates a Latin hypercube sample, based on different optimisation schemes (in this case, random, S optimal and maxmin criteria, respectively; see package documentation for more details). Similarly for the `DiceDesign`, there is a comprehensive list of space-filling designs such as `dmaxDesign()`, `lhsDesign()`, `straussDesign()` and `wspDesign()` with input of $n$ and $p$ as before (among extra parameters for some) that implement algorithms that either maximise the entropy, produce a random Latin hypercube sample, based on Strauss process and WSP algorithm, respectively. These designs all output a $n \times p$ matrix with values between 0 and 1.

@@

Factorial experiments offer a challenge in the construction and allocation of the treatment factors. This is reflected in a number of packages that specifically addresses this challenge. These include `DoE.base` [@DoE.base] for full factorial and (regular and irregular) orthogonal array designs via `fac.design()` and `oa.design()`, respectively, `FrF2` [@FrF2] for fractional 2-level factorial designs using `FrF2()`, `FrF2Large()` or `pb()` Plankett-Burnman designs [@Plackett1946-ly], `conf.design` [@conf.design] for symmetric confounded factorial designs via `conf.design()`, and `BHH2` [@BHH2] to generate full or fractional 2-level factorial design matrix via `ffDesMatrix()`. All these designs generally requires the user to input the number of factors and if the design is allowed to vary in the number of levels, the input of the number of levels for each factor. The construction of some of these designs are catalogue-based, e.g. `oa.design()` in `DoE.base` and the entire package `FrF2.catlg128` [@FrF2.catlg128] is a list of fractional 2-level factorial designs.

@@

@@

This paper uses the `targets` framework [@targets] and `renv` [@renv] for reproducibility, `knitr` [@knitr] and `rmarkdown` [@rmarkdown] for creating reproducible documents, `ggplot2` [@ggplot2], `ggraph` [@ggraph], `ggwordcloud` [@ggwordcloud] and `colorspace` [@colorspace] for visualisation, `kableExtra` [@kableExtra] for customising the table in the Supplementary Material and `tidyverse` [@Wickham2019-mj], `tidytext` [@tidytext], `pluralize` [@pluralize], and `ineq` [@ineq] for data processing and manipulation, and `cranlogs` [@cranlogs] and `ctv` [@ctv] for extracting data. All code to reproduce this paper is found at `r url("https://github.com/emitanaka/paper-DoE-review")`.

>>>>

---
title: Current state and prospects of R-packages for the design of experiments
author:
  - name: Emi Tanaka
    affiliation: Monash University
  - name: Dewi Amaliah 
    affiliation: Monash University
abstract: >
    Re-running an experiment is generally costly and in some cases impossible due to limited resources, so the design of an experiment plays a critical role in increasing the quality of experimental data.  In this article we describe the current state of the R-packages for the design of experiments through an exploratory data analysis of package downloads, package metadata, and the comparison of characteristics with other topics. We observe that experimental designs in practice appear to be sufficiently manufactured by a small number of packages and the development of experimental designs occur in silos. We discuss also the interface designs of widely utilised R packages in the field of experimental design and discuss its future prospects to advance the field in practice.
bibliography: paper.bib
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
keywords: ["experimental design", "CRAN task view", "interface design"]
output: 
  pagedown::html_paged:
    template: assets/html-template.html
    csl: assets/html-template.csl
    self_contained: false
    css: ["assets/html-template.css"]
---

```{r setup, cache = FALSE, include = FALSE}
library(targets)
library(glue)
library(ggwordcloud)
library(tidyverse)
library(patchwork)
library(tidygraph)
library(ggraph)
source("R/plot.R")
knitr::opts_chunk$set(echo = FALSE, 
                      cache = TRUE,
                      cache.path = "cache/",
                      fig.align = 'center', 
                      fig.pos = 'htbp', 
                      fig.path = "figures/",
                      fig.width = 6,
                      fig.height = 6.5,
                      message = FALSE,
                      warning = FALSE)

@@

tar_load(npkgs)
tar_load(gini_yearly_doe)
tar_load(gini_yearly_ctv)
```

@@

The critical role of data collection is well captured in the expression "garbage in, garbage out" -- in other words, if the collected data is rubbish then no analysis, however complex it may be, can make something out of it. A carefully crafted data collection scheme is therefore critical to optimise the information from data. The field of experimental designs is specifically devoted to planning the collection of experimental data largely based on the founding principles by @Fisher1935-qc or an optimisation framework like those described in @Pukelsheim2006-sv. These experimental designs are often constructed with the aid of a statistical software, such as R [@R], so the usage of experimental design software can inform us about some aspects of experimental designs in practice.

@@

Methods for data collection can be dichotomised by the type of data collected -- namely, experimental or observational -- or alternatively, categorised as experimental design (including quasi-experimental design) or survey design. This dichotomisation, to a great extent, is seen in the [Comprehensive R Archive Network (CRAN) task views](https://cran.r-project.org/web/views/) (a volunteer maintained list of R-packages by topic) where R-packages for experimental design are in `r ctv("ExperimentalDesign")` task view and R-packages for survey designs are in `r ctv("OfficialStatistics")` task view. The full list of available topics can be see in Table S1 in the Supplementary Materials. A subset of experimental designs are segregated into `r ctv("ClinicalTrials")` task view, where the focus is on clinical trials with primary interest in sample size calculations. This paper focuses on the packages in `r ctv("ExperimentalDesign")` task view, henceforth referred to as "DoE packages".

@@

A utility of software can also be described by its design to facilitate clear expression and interpretation of the desired experimental design. Certain programming language design can hinder or discourage development of a reliable program [@Wasserman1975-xr].  The immense popularity of `tidyverse` [a collection of R-packages for various stages of data analysis that place enormous emphasis on the interface design by @Wickham2019-mj] is a testament to the impact an interface design can have in practice. The practice of experimental design could be advanced by adopting similar interface design principles across DoE packages.

The paper is organised as follows. Section `r ref("data")` briefly describes the data source used for the analysis in Section `r ref("eda")`; Section `r ref("eda")` presents some insights into the state of the current DoE packages by exploratory data analysis of package download data, text descriptions and comparisons with other CRAN task views; Section `r ref("design")` discuss the interface designs of widely used DoE packages, and we conclude with a discussion in Section `r ref("discussion")` of future prospects in the software development of experimental designs.

@@

To study the DoE packages, we analyse data using three sources of data as described below.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers located across the world that store mirrored versions of R and R-packages. The most popular network is the RStudio mirror (the default server for those that use the RStudio IDE). The RStudio mirror is also the only server that provides a comprehensive daily download logs of R and R-packages since October 2012. The summary data can be easily accessed with the `cranlogs` package [@cranlogs]. This paper uses the data from the beginning of 2013 to end of 2021 (a total of 9 years) for the packages in the CRAN task views.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN relevant to the corresponding topic. There are a total `r n_distinct(gini_yearly_ctv$ctv)` CRAN task views. Table S1 in the Supplementary Materials list the full available topics from the `ctv` package [@ctv].
The list of packages in each CRAN task views (as of `r as.Date(file.info("_targets/objects/ctv_pkgs_data")$mtime)`) are used to contrast the characteristics of DoE packages in Section `r ref("silo")`.

@@

In this section, we derive some conjecture based on the analysis of the data described in Section `r ref("data")`. All results presented are from the exploratory data analysis of observational data, consequently, all interpretations are somewhat speculative and may not be indicative of the true state of the field of experimental design. In particular, any analysis over time are confounded by the fact that the nature of users and package management have also changed over the years. It should be noted that some DoE packages may have been archived or removed from the task view over the years so any cross-sectional analysis presented may not reflect the set of all DoE packages at that particular time period (although we assume such incidences are low).

A subset of DoE packages is not primarily about design of experiments but about the analysis of experimental data. A complete delineation of these packages is difficult as there is almost always at least one function that can aid decisions or constructions of experimental designs (and any categorisation is prone to our subjective bias) so we opted not to remove any DoE packages in the analysis.

## Small, but diverse, set of packages are sufficient for most experimental designs in practice {#popular}

@@

```{r plot-lorenz, fig.height = 4, fig.width = 4, fig.cap = "Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality. The green region shows the area under the Lorenz curve and the red region shows the area of gap in equality."}
tar_read(plot_lorenz2021) +
  plot_setup() +
  labs(y = "Cumulative share of downloads (%)",
       x = "Percentage of DoE packages") +
  theme(plot.margin = margin(r = 20, t = 20)) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10))
```

 If we consider package downloads as a measure of "wealth", then we can consider using the Gini index [@Gini1921-mf] as a measure of download inequality across packages. The ratio of the red region over the total colored regions in Figure `r ref("fig:plot-lorenz")` corresponds to the Gini index for 2021. A Gini index of 0\% indicates equality in downloads across packages while a value of 100\% indicates maximal inequality (all downloads are due to one package). In Figure `r ref("fig:download-share")`, we see that the distributions of the package downloads each year have a heavy right tail with the Gini index ranging from `r scales::percent(min(gini_yearly_doe$gini), 0.1)` to `r scales::percent(max(gini_yearly_doe$gini), 0.1)` across the years 2013 to 2021, indicating that there is a high level of inequality of package downloads, particularly with more pronounced inequality in the last 6 years.

```{r download-share, fig.width = 5.5, fig.height = 4, fig.cap = "The distribution of the number of downloads for each DoE package by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 6 years, the Gini index is consistent above 60\\% each year indicating that most downloads are due to a relatively small number of packages. "}
tar_read(plot_download_dist) +
  labs(y = "Number of downloads")
```

The increase in the number of packages that are not highly downloaded may mean that **_there are more packages to construct niche experimental designs_**. Some examples of these packages include `qtlDesign`, `PwrGSD` and `Crossover` made for QTL experiments, group sequential designs and crossover trials, respectively. These packages would naturally have a smaller number of potential users. Counterfactual to this, the increase could be due to other external factors, such as an increase in the number of skilled developers (and thus more pacakge contributions), a change in CRAN policy or management to add packages (either to CRAN and/or task view), and/or that new packages are still yet to amass users. While there is an argument that low download counts are due to the low utility and/or quality of the packages, packages in CRAN task views are selected by expert maintainers, thus we can reasonably assume that any package listed in CRAN task view are of a decent utility and quality.

If the downloads are reflective of the experimental designs used in practice, **_small set of packages appear to be sufficient for most to construct the full set of designs of experiments needed in practice_**. Packages of course evolve and the top downloaded packages have had regular updates that may have broadened its scope from its previous releases.

While in absolute terms the Gini index is high for `r ctv("ExperimentalDesign")` task view (`r scales::percent(min(gini_yearly_doe$gini), 0.1)` to `r scales::percent(max(gini_yearly_doe$gini), 0.1)`), the inequality is not as severe as other CRAN task views as shown in Figure `r ref("fig:fig-gini-all-ctvs")`. We can see in Figure `r ref("fig:fig-gini-all-ctvs")` that the Gini index is generally increasing for DoE packages (as is generally the case for other CRAN task views as shown in Figure S1 in the Supplementary Materials) but most other CRAN task views have a Gini index of over 75%. This suggests that other CRAN task views may have dominant standards and in comparison to other topics, there are more **_diverse approaches to designing experiments_** and thus no single DoE package is dominant. This observation however doesn't take into account other approaches to generating experimental designs, like the proprietary software, CycDesignN [@cycdesignn] and SAS [@sas1985sas], which may be widely used.

```{r fig-gini-all-ctvs, fig.height = 4, fig.width = 4, fig.cap = "The points show the Gini index of the download counts by year for ExperimentalDesign task view with the color showing the number of packages. The grey line shows the distribution of the Gini index across years for all other CRAN task views. See Figure S1 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view."}
gini_yearly_ctv %>% 
  ungroup() %>%  
  arrange(ctv, year) %>% 
  mutate(ctv = fct_reorder(ctv, gini, last),
         year = factor(year),
         gini = 100 * gini) %>%  
  ggplot(aes(year, gini, group = ctv)) + 
  geom_line(color = "grey") +
  geom_line(data = ~filter(.x, ctv=="ExperimentalDesign")) + 
  geom_point(data = ~filter(.x, ctv=="ExperimentalDesign"),
             aes(color = n))  +
  plot_setup() +
  labs(x = "Year", y = "Gini index (%)", color = "# of packages",
       title = "ExperimentalDesign CRAN task view") +
  guides(color = guide_colorbar(direction = "horizontal")) +
  theme(legend.position = "bottom") +
  scale_colour_distiller(palette = "YlOrRd", direction = 1) 
```

## There is a lack of adaptation of new or innovative experimental designs  {#ranking}

@@

```{r rank-over-time, fig.width = 5.5, fig.height = 4.5, fig.cap = "The plot shows the rank of top 10 downloaded packages by year. Packages that do not appear in top 10 for at least two periods are omitted from the plot. Most packages are consistently in the top 10 for the period shown."}
tar_read(plot_rank_dist) + labs(color = "Package")
```

@@

```{r established-packages, fig.height = 5, fig.cap = "The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log (with base 10) of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R."}
tar_read(cran_rank_yearly) %>%
  mutate(nupdates = map2_int(package, year, ~{
    tar_read(updates_all) %>% 
      filter(package == .x,
             lubridate::year(update) <= .y) %>% 
      nrow()
  })) %>% 
  ggplot(aes(first, total/1000)) +
  geom_point(aes(color = nupdates)) + 
  geom_label(data = ~ .x %>% 
              group_by(year) %>% 
              summarise(R = cor(as.numeric(first), log10(total/1000)) %>% 
                          scales::comma(0.01) %>% 
                          paste0("r = ", .),
                        y = quantile(total/1000, 0.99)),
            aes(label = R, y = y, x = as.Date("2014-11-01")),
            size = 3) +
  facet_wrap(~year, scale = "free_y") +
  geom_smooth(se = FALSE, 
              method = lm, 
              formula = y ~ x) +
  scale_y_log10() + 
  scale_color_continuous(trans = "log10") + 
  plot_setup() +
  labs(x = "First release date",
       y = "Total download (1000s)",
       color = "# of updates") +
  theme(legend.position = "bottom")
```

@@

```{r wordcloud-over-time, fig.width = 5.5, fig.height = 5, fig.cap = "The above figure shows the word cloud of bigrams from the title and descriptions of the DoE packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram."}
tar_read(plot_wordcloud2) + 
  scale_size_area(max_size = 5) +
  theme(legend.text = element_text(size = 10),
        legend.title = element_text(size = 12)) +
  labs(color = "Total downloads (%)")
```

@@

In this section, we discuss these interface designs broadly, with examples illustrated from the top downloaded packages (shown in Figure `r ref("fig:rank-over-time")`). We exclude `ez` and `DiceKriging` from the discussion as the former  is predominately visualisation of experimental data and latter is about the analysis of computer experiments in addition to belonging to the same suite of packages as `DiceDesign` [@DiceDesign].

@@

Computer experiments, as discussed in Section `r ref("topics")`, which generally involve space filling designs, have been on the rise. The exemplar for this type of designs are `lhs` [@lhs] and `DiceDesign`. For the `lhs` package, functions like `randomLHS()`, `optimumLHS()`, and `maximinLHS()`, require users to specify the sample size ($n$) and the number of variables ($p$) then it generates a Latin hypercube sample, based on different optimisation schemes (in this case, random, S optimal and maxmin criteria, respectively; see package documentation for more details). Similarly for the `DiceDesign`, there is a comprehensive list of space-filling designs such as `dmaxDesign()`, `lhsDesign()`, `straussDesign()` and `wspDesign()` with input of $n$ and $p$ as before (among extra parameters for some) that implement algorithms that either maximise the entropy, produce a random Latin hypercube sample, based on Strauss process and WSP algorithm, respectively. These designs all output a $n \times p$ matrix with values between 0 and 1.

@@

Factorial experiments offer a challenge in the construction and allocation of the treatment factors. This is reflected in a number of packages that specifically addresses this challenge. These include `DoE.base` [@DoE.base] for full factorial and (regular and irregular) orthogonal array designs via `fac.design()` and `oa.design()`, respectively, `FrF2` [@FrF2] for fractional 2-level factorial designs using `FrF2()`, `FrF2Large()` or `pb()` Plankett-Burnman designs [@Plackett1946-ly], `conf.design` [@conf.design] for symmetric confounded factorial designs via `conf.design()`, and `BHH2` [@BHH2] to generate full or fractional 2-level factorial design matrix via `ffDesMatrix()`. All these designs generally requires the user to input the number of factors and if the design is allowed to vary in the number of levels, the input of the number of levels for each factor. The construction of some of these designs are catalogue-based, e.g. `oa.design()` in `DoE.base`.

@@

`r if(knitr::is_latex_output()) '<!--'`

# Supplemental materials

The supplementary materials can be found [here](supp.pdf).

`r if(knitr::is_latex_output()) '-->'`

@@

This paper uses `targets` [@targets] and `renv` [@renv] for reproducibility, `knitr` [@knitr] and `rmarkdown` [@rmarkdown] for creating reproducible documents, `ggplot2` [@ggplot2], `ggraph` [@ggraph], `ggwordcloud` [@ggwordcloud] and `colorspace` [@colorspace] for visualisation, `kableExtra` [@kableExtra] for customising the table in the Supplementary Material and `tidyverse` [@Wickham2019-mj], `tidytext` [@tidytext], `pluralize` [@pluralize], and `ineq` [@ineq] for data processing and manipulation, and `cranlogs` [@cranlogs] and `ctv` [@ctv] for extracting data. All code to reproduce this paper is found at `r url("https://github.com/emitanaka/paper-DoE-review")`.
